Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-29 paper-1/manuscript.tex
Editing

Thus, we have finally traced the greater response to A in the dev sequence to its roots, and revealed two complementary roles of TA: The frequent presentation of non-target stimulus B in the dev sequence and concomitant increased activity in a subset of neurons responding early to B causes \emph{local suppression} by TA in this subset, leading to smaller network responses. By contrast, stimuli in the msc sequence are presented more sporadically, allowing local suppression to decay, thus leading to larger network responses. This difference in non-target response sizes -- small in dev, large in msc -- is then reflected in network-wide \emph{global suppression}, such that the network is more excitable to stimulation with A (or other non-standard inputs) in the dev sequence.

@@

To understand how STD affects deviance detection, we first examined the effects of TA in the full model, following the analysis on the ablated model laid out in the previous section. Starting at the cause of the reduced response in non-target trials, we show in Figure \ref{fig7}a the full model equivalent of Figure \ref{fig6}c and d. The contrasts (dev - msc) in the non-target response magnitude and in the resulting $V_{TA}$ levels at the start of B trials both retained a clear separation between the portion of neurons responding early, which showed increased responses (T = 5.44e+03, p = 1.02e-06; one-sided Wilcoxon, n = 120) and increased TA levels (T = 3.41e+03, p = 9.51e-08), and the neurons responding later, whose response (T = 282, p = 9.19e-19) and TA levels (T = 72, p = 2.49e-17) decreased in the dev sequence. Likewise, the correlation across excitatory neurons between non-target response contrast and the contrast in $V_{TA}$ before B trials remained positive (T = 7.26e+03, p = 9.86e-22) and very strong (median Pearson's $\rho$ = 0.992) This indicates that local suppression to B mediated by TA remained a key driver of the reduced response to non-target trials.

Inspecting the non-target response reduction in the full model in Figure \ref{fig7}b, we found that it was on average larger than in the ablated model (T = 4.16e+03, p = 0.0175; one-sided Wilcoxon, n = 120), with the grand medians differing by 36\%. Based on the ablated model analysis in Figure \ref{fig5}, we expected reduced non-target responses to similarly reduce the TA levels at the start of A trials. Indeed, the non-target response contrast remained tightly correlated with the contrast (dev - msc) in $V_{TA}$ at the start of target trials (median Pearson's $\rho$ = 0.989), as shown in Figure \ref{fig7}c. Correspondingly, $V_{TA}$ was reduced in dev, as shown in Figure \ref{fig7}d, and this too by a slighly greater amount in the full model than in the ablated model (T = 4.37e+03, p = 0.0257).

Finally, based on the ablated model analysis in Figure \ref{fig4}, we sought to confirm that lowered $V_{TA}$ before target trials was responsible for the increased target response. As shown in Figure \ref{fig7}e, the contrasts (dev - msc) in $V_{TA}$ at the start of target trials and in response magnitude to A were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response was clearly increased in dev over msc, as shown in Figure \ref{fig7}f, and this increase was greater than in the ablated model (T = 1.94e+03, p = 0.000723), with the grand medians differing by 94\%. This greater contrast, of course, directly reflects the greater DDI reported in Figure \ref{fig3}.

@@

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than the non-target msc responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences. This would mean that STD is capable only of local suppression, reducing the standard response, but not of global suppression, leaving networks with only STD unable to differentiate between dev and msc target trials.

We had good reasons for this assumption: Firstly, referring back to Figure \ref{fig3}, notice that the ablated model with STD only produced only a positive RSI (i.e., the target response in msc was greater than in std), but not a significant DDI (i.e., no difference between target dev and msc responses). Secondly, we note that updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B could lead to a build-up of depression strong enough to suppress its own response.
To show that the network does not carry information beyond the most recent trial in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial.

To show that STD being carried over between repeated trials was responsible for the reduction of non-target dev pulses relative to msc, however, we needed to estimate the extent to which activity of a given neuron (or, alternatively, its membrane potential $V_m$) was affected by changes in STD. Unlike $V_{TA}$, which directly affects the neuron to which the value belongs, $x_j$ exerts its effects on the other side of a synapse. Therefore, we devised a measure $D$ to estimate the postsynaptic impact of STD in terms of membrane potential lost, mirroring the effect of $V_{TA}$ in the sense that decreasing $V_m$ and increasing the firing threshold both increase the depolarization necessary to make a neuron fire.
A measure of STD is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. Here, we used the average trial response $R_j^S$ of neuron $j$ to a given stimulus $S$, in number of spikes, as the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at rest (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $R_j^S k x_j^S$. For this estimate, we ignored added depression due to repeated spikes within a trial, instead defining $x_j^S$ as the average depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in a given neuron $i$ as
\begin{equation}
    D_i^S = \sum_{j \in pre} R_j^S k (1-x_j^S)
\end{equation}
Notice that $D$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $V_{TA}$, albeit more approximate. Lastly, since we contrasted this measure between conditions (see below), we aimed to avoid effects purely due to differences in presynaptic activity. Therefore, we calculated $R^S$ as the average response across $S$ trials in all conditions, leaving the contrast entirely in the hands of $x^S$.

@@

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target dev trials, relative to msc, whereas differences between target dev and msc trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across dev and msc conditions as with $V_{TA}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both conditions. The msc sequence, however, clearly evoked greater depression, as borne out in the contrast (dev - msc), which was negative throughout. Conversely, depression in B trials (bottom) was spatially more clustered around the stimulation site in both conditions. In the msc sequence, we can attribute the difference between A and B primarily to the response pattern: The context was the same, but responses to B tended not to spread out as much in this network. However, in the dev sequence, the B response was limited almost exclusively to a very local patch around the stimulation site, with depression therefore similarly limited. As a result, the contrast (dev - msc) showed slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicating greatly increased synaptic depression in dev.

@@

To show that increased local suppression by STD reduces the non-target response, we correlated the median contrast in $D^B$ among the early portion of networks with the median contrast of the total response to B. As we show in Figure \ref{fig8}d, higher $\Delta D^B$ was clearly predictive of a lower network response ($\rho$ = -0.581). This confirms again the integrity of the measure $D$, but also that STD-mediated local suppression played a role in reducing the non-target response in the dev sequence. % Possible concern: The equivalent hasn't been shown in TA (Fig6), but likely could be (in the same fashion; dTA ~ dXB)

Finally, given the very clear contrast in $D^A$ in the sample network, we considered the possibility that, in the context of the full model, STD could also evoke global suppression in the msc sequence, contrary to our stated hypothesis. If so, we should expect to see less $D^A$ in dev than msc. To confirm this, we analysed the contrast (dev - msc) in $D^A$ across networks and stimuli, and found its network median to be robustly negative (T = 955, p = 1.23e-12; one-sided Wilcoxon, n = 120). In addition, as shown in Figure \ref{fig8}e, we correlated the contrast in $D^A$ with the target response contrast in each network, and found that weaker depression in dev predicted a greater response in most networks (T = 724, p = 1.37e-14), though the median correlation coefficient $\rho$ was only -0.282. We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see Figure \ref{sfig8}). Taken together, we cautiously conclude that, while there is clear evidence for greater global suppression in msc than dev in terms of $D$, this difference appears to manifest weakly in activity differences.

To sum up, in the full model, STD acted in much the same way as TA, causing both a reduction of the response to the frequently presented standard stimuli (dev B) by local suppression, and an increase in the response to deviant stimuli (dev A) by relief from global suppression, relative to control. Yet, as we showed in Figure \ref{fig7} and in the previous section, a decrease in dev B responses also entailed lower $V_{TA}$, and higher dev A responses by consequence. Thus, adding STD to the model increased the DDI both directly, by reducing global suppression, and indirectly, by altering the pattern of TA by way of local suppression of B.

We note that the direct and indirect effect of STD on global suppression appear to be on the same order of magnitude. To see this, compare the change in the $V_{TA}$ contrast at the start of A trials as a result of adding STD to the model (Figure \ref{fig7}d, bottom) with the corresponding $\Delta D^A$ (Figure \ref{fig8}e), both of which are on the order of $\pm$ 0.1 mV. Conversely, notice that, since TA and STD work in similar fashion to increase the dev target response, it is very plausible that TA likewise increases the effect of STD, by enhancing the STD-mediated local suppression of B and thereby increasing the contrast in STD-mediated global suppression.

>>>>

Thus, we have finally traced the greater response to A in the dev sequence to its roots, and revealed two complementary roles of TA: The frequent presentation of non-target stimulus B in the dev sequence and concomitant increased activity in a subset of neurons responding early to B causes \emph{local suppression} by TA in this subset, leading to smaller network responses. By contrast, stimuli in the msc sequence are presented more sporadically, allowing local suppression to decay, thus leading to larger network responses. This difference in non-target response sizes -- small in dev, large in msc -- is then reflected in network-wide \emph{global suppression}, such that the network is more excitable to stimulation with A (or other non-standard inputs) in the dev sequence than in the control context.

@@

To understand how STD affects deviance detection and causes this difference in response reduction, we first examined the effects of TA in the full model, following the analysis on the ablated model laid out in the previous section. Starting at the cause of the reduced response in non-target trials, we show in Figure \ref{fig7}a the full model equivalent of Figure \ref{fig6}c and d. The contrasts (dev - msc) in the non-target response magnitude and in the resulting $V_{TA}$ levels at the start of B trials both retained a clear separation between the portion of neurons responding early, which showed increased responses (T = 5.44e+03, p = 1.02e-06; one-sided Wilcoxon, n = 120) and increased TA levels (T = 3.41e+03, p = 9.51e-08), and the neurons responding later, whose response (T = 282, p = 9.19e-19) and TA levels (T = 72, p = 2.49e-17) decreased in the dev sequence. Likewise, the correlation across excitatory neurons between these two contrasts remained positive (T = 7.26e+03, p = 9.86e-22) and very strong (median Pearson's $\rho$ = 0.992). This indicates that local suppression to B mediated by TA remained a key driver of the reduced response to non-target trials.

Comparing the non-target response reductions in Figure \ref{fig7}b, we found that they were on average larger in the full model than in the ablated model (T = 4.16e+03, p = 0.0175; one-sided Wilcoxon, n = 120), with the grand medians differing by 36\%. Based on the ablated model analysis in Figure \ref{fig5}, we expected reduced non-target responses to similarly reduce the TA levels at the start of A trials. Indeed, the non-target response contrast remained tightly correlated with the contrast (dev - msc) in $V_{TA}$ at the start of target trials (median Pearson's $\rho$ = 0.989), as shown in Figure \ref{fig7}c. Correspondingly, $V_{TA}$ was reduced in dev, as shown in Figure \ref{fig7}d, and this too by a slighly greater amount in the full model than in the ablated model (T = 4.37e+03, p = 0.0257).

Finally, based on the ablated model analysis in Figure \ref{fig4}, we sought to confirm that lowered $V_{TA}$ before target trials was responsible for the increased target response. As shown in Figure \ref{fig7}e, the contrasts (dev - msc) in $V_{TA}$ at the start of target trials and in response magnitude to A were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response was clearly increased in dev over msc, as shown in Figure \ref{fig7}f, and this increase was greater in the full model than in the ablated model (T = 1.94e+03, p = 0.000723), with the grand medians differing by 94\%. This greater contrast, of course, directly reflects the greater DDI reported in Figure \ref{fig3}.

@@

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than non-target msc responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences. This would mean that STD is capable only of local suppression, reducing the standard response, but not of global suppression, leaving networks with only STD unable to differentiate between dev and msc target trials.

We had good reasons for this assumption: Firstly, referring back to Figure \ref{fig3}, notice that the ablated model with STD only produced a positive RSI (i.e., the target response in msc was greater than in std), but not a significant DDI (i.e., no difference between target dev and msc responses). Secondly, we note that updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B may have been able to lead to a build-up of depression strong enough to suppress its own response.

To show in the sample network that the model does not maintain a history beyond the most recent trial, we correlated the number of spikes in dev target trials with $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was strong and very clearly positive in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; Wald test, n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial.

To show that STD being carried over between repeated trials was responsible for the reduction of non-target dev pulses relative to msc, however, we needed to estimate the extent to which activity of a given neuron (or, alternatively, its membrane potential $V_m$) was affected by changes in STD. Unlike $V_{TA}$, which directly affects the neuron to which the value belongs, $x_j$ exerts its effects on the other side of a synapse. Therefore, we devised a measure $D$ to estimate the postsynaptic impact of STD in terms of membrane potential lost, mirroring the effect of $V_{TA}$ in the sense that decreasing $V_m$ and increasing the firing threshold both increase the depolarization necessary to make a neuron fire.

A measure of STD is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. Here, we used the average trial response $R_j^S$ of neuron $j$ to a given stimulus $S$, in number of spikes, as a measure of the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at $V_m = V_{rest}$ (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $R_j^S k x_j^S$. For this estimate, we ignored added depression due to repeated spikes within a trial, instead defining $x_j^S$ as the mean depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in a given neuron $i$ as
\begin{equation}
    D_i^S = \sum_{j \in pre} R_j^S k (1-x_j^S)
\end{equation}
Notice that $D$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $V_{TA}$, albeit more approximate. Lastly, since we contrasted this measure between conditions (see below), we aimed to avoid effects purely due to differences in presynaptic activity. Therefore, we calculated $R^S$ as the average response across $S$ trials in all conditions, leaving the contrast entirely in the hands of $x^S$.

@@

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target dev trials, relative to msc, whereas differences between target dev and msc trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across dev and msc conditions as with $V_{TA}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both conditions. The msc sequence, however, clearly evoked greater depression, as borne out in the contrast (dev - msc), which was negative throughout. Conversely, depression in B trials (bottom) was spatially more clustered around the stimulation site in both conditions. In the msc sequence, we can attribute the difference between A and B primarily to different response patterns: The context, and thus $x$, was approximately the same, but responses to B tended not to spread out as much in this network. However, in the dev sequence, the B response was limited almost exclusively to a very local patch around the stimulation site, with depression therefore similarly spatially limited. As a result, the contrast (dev - msc) showed slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicating greatly increased synaptic depression in dev.

@@

To show that increased local suppression by STD reduces the non-target response, we correlated the median contrast in $D^B$ among the early portion of networks with the median contrast of the total response to B, echoing Figure \ref{fig6}e. As we show in Figure \ref{fig8}d, higher $\Delta D^B$ was clearly predictive of a lower network response ($\rho$ = -0.581, p = 1.81e-12; Wald test, n = 1000). This confirms again the integrity of the measure $D$, but also that STD-mediated local suppression played a role in reducing the non-target response in the dev sequence.

Finally, given the very clear contrast in $D^A$ in the sample network, we considered the possibility that, in the context of the full model, STD could also evoke global suppression in the msc sequence, contrary to our stated hypothesis. If so, we should expect to see less $D^A$ in dev than msc. To confirm this, we analysed the contrast (dev - msc) in $D^A$ across networks and stimuli, and found its network median to be robustly negative (T = 955, p = 1.23e-12; one-sided Wilcoxon, n = 120). In addition, as shown in Figure \ref{fig8}e, we correlated the contrast in $D^A$ with the target response contrast across neurons in each network, and found that weaker depression in dev predicted a greater response in most networks (T = 724, p = 1.37e-14), though the median correlation coefficient $\rho$ was only -0.282. We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see Figure \ref{sfig8}). Taken together, we cautiously conclude that, while there is clear evidence for greater global suppression in msc than dev in terms of $D$, this difference appears to manifest only weakly in activity differences.

To sum up, in the full model, STD acted in much the same way as TA, causing both a reduction of the response to the frequently presented standard stimuli (dev B) by local suppression, and an increase in the response to deviant stimuli (dev A) by relief from global suppression, relative to control. Yet, as we showed in Figure \ref{fig7} and in the previous section, a decrease in dev B responses also entailed lower $V_{TA}$, which independently lead to higher dev A responses. Thus, adding STD to the model increased the DDI both directly, through its intrinsic effects in local and global suppression, and indirectly, by altering the pattern of TA by way of local suppression of B.

We note that the direct and indirect effects of STD on global suppression appear to be on the same order of magnitude. To see this, compare the change in the $V_{TA}$ contrast at the start of A trials as a result of adding STD to the model (Figure \ref{fig7}d, bottom) with the corresponding $\Delta D^A$ (Figure \ref{fig8}e), both of which are on the order of $\approx$ 0.1 mV. Conversely, notice that, since TA and STD work in similar fashion to increase the dev target response, it is very plausible that TA likewise increases the effect of STD, by enhancing the STD-mediated local suppression of B and thereby increasing the contrast in STD-mediated global suppression. Thus, this synergy allows the individually weak STD to exert a significantly larger effect when paired with the stronger TA.
